---
title: "Gagan Makhiza"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reproducible research Course Project 1


## 1. Loading and preprocessing the data
###1.1. Unzip and read the data from the GitHub repository. Dimensional exploration.

1. Load the data (i.e. read.csv()).

```{r}
act <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day.

```{r}
library(dplyr)
actSum <- act %>% group_by(date) %>% summarise(stepSum = sum(steps))
```

2. Make a histogram of the total number of steps taken each day.

```{r}
hist(actSum$stepSum, breaks = 10, main = "total number of steps taken each day", xlab = "Steps", col = "wheat")
```

3. Calculate and report the mean and median of the total number of steps taken per day.

```{r}
actMean <- act %>% group_by(date) %>% summarise(stepSum = mean(steps))
actMedian <- act %>% group_by(date) %>% summarise(stepSum = median(steps))
```

## What is the average daily activity pattern?

1. a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r}
actInt <- act %>% group_by(interval) %>% summarise(stepInt = sum(steps, na.rm = T))
plot(actInt$interval, actInt$stepInt, type = "l", main = "average no. of steps taken", ylab = "averaged across all day", xlab = "5-min interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxint <- actInt[actInt$stepInt == max(actInt), 1]
maxint
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset.

```{r}
sum(is.na(act))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Strategy taken is to substitute each NA with a fixed value. I set the fixed value equivalent to the overall mean of the variable act$steps.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
act2 <- act
act2$steps[is.na(act2$steps)] <- mean(na.omit(act$steps))
sum(is.na(act2))
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
## Report

sum_mean_median_day <- group_by(act2, date) %>%
     summarise(sum_steps = sum(steps),
               mean_steps = mean(steps),
               median_steps = median(steps))
 sum_mean_median_day
 
 ## Histogram
 par(mfcol = c(2,1))
 hist(sum_mean_median_day$sum_steps, main = "Histogram of the total steps per day AFTER missing data filling", xlab = "Steps", breaks = 16, col = "wheat")
 hist(actSum$stepSum, main = "Histogram of the total steps per day BEFORE missing data filling", xlab = "Steps",   breaks = 16, col = "wheat", ylim = c(0,15))
```

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
act2$date = as.Date(act2$date, "%Y-%m-%d")
state <- ifelse(weekdays(act2$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
state <- as.factor(state)
act2$state <- state
```

2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
library(ggplot2)
average_steps_state <- group_by(act2, state, interval) %>% summarise(average_steps = mean(steps))

ggplot(average_steps_state, aes(x = interval, y = average_steps)) +
    geom_line() + 
    facet_grid(state ~ .) + 
    labs(x = "Interval") + 
    labs(y = "Number of steps")

```
